---
title: "Career MAP Program: Draft Analysis Report"
author: "Blondie Beatrice Ndebele"
date: "November 2, 2025"
format:
  pdf: default
---

# 

```{r}
#libraries
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(tidyr)
library(car)
library(janitor)
library(caret)
library(glmnet)
library(tigris)
library(sf)
library(httr)
library(jsonlite)
library(xgboost)
library(Matrix)
library(randomForest)
library(purrr)
library(cluster)
library(factoextra)
```

## Data Loading and Cleaning

```{r}
# Load all sheets
df <- "Rents & Relocations_AU.xlsx"
sheets <- excel_sheets(df)
sheets_short <- c("old", "new", "curr")


rent <- map(sheets, function(sh) read_excel(df, sheet = sh)) |> 
  set_names(nm = sheets_short)
```

```{r}
#View all sheets
head(rent)
```

## Original Unit Rent

**Variables of interest: Response ID, Original Unit Rent and Original Zip code**

```{r}
# Access first sheet
old_rents <- rent[['old']]
head(old_rents)

#Check Structure 
glimpse(old_rents)
```

```{r}
cat("Missing values present:", any(is.na(old_rents)), "\n")
```

## Original Unit Rent Cleaning and Summary

```{r}
#change names to snake case 
old_rents <- clean_names(old_rents)

# Check new names
colnames(old_rents)
```

```{r}
#Extracting only 5 digits of the zipcode 
old_rents <- old_rents |> 
  mutate(original_zip_code = str_extract(old_rents$original_zip_code, 
                                         "\\d{5}$"))

head(old_rents)
```

```{r}
# View clean summary of the old rent data
skim(old_rents)
```

```{r}
#Selecting variables of interest
old_rents<- old_rents |>
  select(response_id, original_unit_rent, original_zip_code) |>
  filter(!is.na(original_zip_code)) |> 
  mutate(original_unit_rent = as.numeric(original_unit_rent)) # Convert to numeric
head(old_rents)
```

```{r}
# Descriptive Statistics 
old_rents |>
  summarize(
    mean = mean(original_unit_rent, na.rm = TRUE),
    median = median(original_unit_rent, na.rm = TRUE),
    min = min(original_unit_rent, na.rm = TRUE),
    max = max(original_unit_rent, na.rm = TRUE),
    sd = sd(original_unit_rent, na.rm = TRUE)
  )
```

## Relocation Rents 
```{r}
#EDA to understand structure of second sheet(new_rents)
new_rents <- rent[["new"]]
head(new_rents)

new_rents
```

```{r}
cat("Missing values present:", any(is.na(new_rents)), "\n")
```

```{r}
# % missing values per column
missing_sorted <- sort(colMeans(is.na(new_rents)) * 100, decreasing = TRUE)
head(missing_sorted)
```

```{r}
#Clean names 
new_rents <- clean_names(new_rents)

# Check new names
colnames(new_rents)
```

```{r}
#Selecting variables that are important for the analysis
new_rents <- new_rents |>
  select(response_id, rent_amount, new_zip_code, proposed_new_unit_size, 
         household_size, utilities_included) |>
  mutate(
    new_zip_code = str_extract(new_zip_code, "\\d{5}$"),
    rent_amount = as.numeric(rent_amount),
    new_zip_code = as.factor(new_zip_code),
    proposed_new_unit_size = as.factor(proposed_new_unit_size)
  ) |>
  arrange(desc(rent_amount)) |>
  drop_na()

head(new_rents)
```

```{r}
#Selecting variables of interest and dropping missing values
new_rents <- new_rents |> 
  select(response_id, rent_amount, new_zip_code,proposed_new_unit_size, 
         household_size, utilities_included) |>
  mutate(new_zip_code = str_extract(new_rents$new_zip_code,"\\d{5}$")) |>
  mutate(rent_amount = as.numeric(rent_amount)) |>
  arrange(desc(rent_amount)) |>
  drop_na()
head(new_rents)
```

**New Unit Rents Summary Statistics**

```{r}
# Descriptive Statistics 
rent_stats2 <-new_rents |>
  summarize(
    mean = mean(rent_amount, na.rm = TRUE),
    median = median(rent_amount, na.rm = TRUE),
    min = min(rent_amount, na.rm = TRUE),
    max = max(rent_amount, na.rm = TRUE),
    sd = sd(rent_amount, na.rm = TRUE)
  )
rent_stats2
```

### Side by side visualization of original rents and relocation rents 

```{r, warning=FALSE}
#| fig-width: 10
#| fig-height: 4
#| dpi: 200


# Plot 1: Old Rents Visualization
# Visualization of the original unit rents to understand the structure of the data.
p1 <- old_rents |> 
  filter(!is.na(original_unit_rent)) |> 
  ggplot(aes(x = original_unit_rent)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 30) +
  geom_vline(aes(xintercept = rent_stats$median), color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = rent_stats$median * 1.03, y = Inf, vjust = 1.5, label = paste("Median:", dollar(rent_stats$median)), color = "red", size = 4) +
  labs(title = "Distribution of Original Unit Rents", x = "", y = "") +
  scale_x_continuous(labels = dollar) +
  theme_minimal()

# Plot 2: New rent distribution after relocation
p2 <- new_rents |> 
  filter(!is.na(rent_amount)) |> 
  ggplot(aes(x = rent_amount)) +
  geom_histogram(fill = "pink", color = "black", bins = 30) +
  geom_vline(aes(xintercept = rent_stats2$median), color = "black", linetype = "dashed", size = 1) +
  annotate("text", x = rent_stats2$median * 1.03, y = Inf, vjust = 1.5, label = paste("Median:", dollar(rent_stats2$median)), color = "red", size = 4) +
  labs(title = "Distribution of Unit Rents After Relocation", x = "", y = "") +
  scale_x_continuous(labels = dollar) +
  theme_minimal()

# Compare two plots 
p1 | p2

```



**Number of program clients in each Zip code before and after relocation.**

```{r, include=FALSE}
# Zipcodes by number of program clients to check the distribution of clients in DC zipcodes and the median rent in each zip code. 
zipcode_counts <- old_rents|>
  group_by(original_zip_code) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count))

# Top 5 zipcodes
top_zipcodes <- zipcode_counts |> 
  filter(row_number() <= 5) |> 
  pull(original_zip_code) |> sort()
head(top_zipcodes)
```

```{r, include=FALSE}
# New zipcodes by number of program clients
new_zip <- new_rents|>
  filter(!is.na(new_zip_code)) |>
  group_by(new_zip_code) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count))

# Top 5 new zipcodes
top_new_zip <- new_zip |> 
  filter(row_number() <= 5) |> 
  pull(new_zip_code) |> 
  sort()
  
head(top_new_zip)
```

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 200


# First bar chart: Number of clients by Zip Code
p1 <- ggplot(zipcode_counts, aes(x = reorder(original_zip_code, count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = count), hjust = 1) +
  scale_y_continuous(position = "right") +
  coord_flip() +
  labs(
    title = "Number of clients by Zip Code",
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())

# Second bar chart: Number of relocated clients by Zip Code
p2 <- ggplot(new_zip, aes(x = reorder(new_zip_code, count), y = count)) +
  geom_bar(stat = "identity", fill = "salmon") +
  geom_text(aes(label = count), hjust = 1) +
  scale_y_continuous(position = "right") +
  coord_flip() +
  labs(
    title = "Number of relocated clients by Zip Code",
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())

# Compare number of clients by zipcodes. 
p1 | p2
```

### Average and Median Original Rent by Zip code

```{r}
# Average rent by zipcode
rent_by_zip <- old_rents |>
  group_by(original_zip_code) |>
  summarize(average_rent = mean(original_unit_rent, na.rm = TRUE),
            median_rent = median(original_unit_rent, na.rm = TRUE)) |>
  arrange(average_rent)
rent_by_zip
```

```{r}
# Bar chart of average rent by zipcode
ggplot(rent_by_zip, aes(x = reorder(original_zip_code, average_rent), y = average_rent)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_point(aes(y = median_rent)) +
  geom_text(aes(label = round(average_rent, 0)), hjust = 0.1) + 
  coord_flip() +
  labs(title = "Average unit rent by zipcode",
       x = "Zipcode",
       y = "Average Rent ($)") +
  theme_minimal()
```






## Old Rents Visualization

### Distribution of Original Unit Rent

```{r}
# Distribution Plot of original unit rents
old_rents |> 
  filter(!is.na(original_unit_rent)) |> 
  ggplot(aes(x = original_unit_rent)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 30) +
  labs(title = "Distribution of Original Unit Rents", x = "Rent($)", 
       y = "Count") +
  theme_minimal()
```

Original unit rents are concentrated around 1200- 1500, with most units being below 2000 and a few outliers above 3000

### Zipcodes by number of program clients

```{r}
# Zipcodes by number of program clients
zipcode_counts <- old_rents|>
  group_by(original_zip_code) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count))

# Top 5 zipcodes
top_zipcodes <- zipcode_counts |> 
  filter(row_number() <= 5) |> 
  pull(original_zip_code) |> sort()
head(top_zipcodes)
```

```{r}
#Visualize, bar chart 
ggplot(zipcode_counts, aes(x = reorder(original_zip_code, count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = count), hjust = -0.1) + # Count labels outside bars
  scale_y_continuous(position = "right") +
  coord_flip() +  # Flip bars horizontal
  labs(
    title = "Number of clients by Zip Code",
    subtitle = paste("Most clients are coming from 5 zip codes: ", 
                     paste(top_zipcodes, collapse = ", ")),
       x = "",
       y = ""
    ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

### Average and Median rent by zip code

```{r}
# Average rent by zipcode
rent_by_zip <- old_rents |>
  group_by(original_zip_code) |>
  summarize(average_rent = mean(original_unit_rent, na.rm = TRUE),
            median_rent = median(original_unit_rent, na.rm = TRUE)) |>
  arrange(average_rent)
rent_by_zip
```

```{r}
# Bar chart of average rent by zipcode
ggplot(rent_by_zip, aes(x = reorder(original_zip_code, average_rent), y = average_rent)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_point(aes(y = median_rent)) +
  geom_text(aes(label = round(average_rent, 0)), hjust = 0.1) + 
  coord_flip() +
  labs(title = "Average and Median unit rent by zipcode",
       x = "",
       y = "") +
  theme_minimal()
```

### Average and Median Relocation Rents by Zip code
```{r}
# Average and median rent by zipcode
new_rent_by_zip <- new_rents |>
  filter(!is.na(new_zip_code), !is.na(rent_amount)) |>
  group_by(new_zip_code) |>
  summarize(
    avg_new_rent = mean(rent_amount),
    median_new_rent = median(rent_amount),
    .groups = "drop"
  )
new_rent_by_zip
```

```{r}
# Visualization of average and median new rent by zipcode
ggplot(new_rent_by_zip, aes(x = reorder(new_zip_code, avg_new_rent), y = avg_new_rent)) +
  geom_bar(stat = "identity", fill = "indianred") +
  geom_point(aes(y = median_new_rent)) +
  geom_text(aes(label = round(avg_new_rent, 0)), hjust = 1) + 
  coord_flip() +
  labs(title = "Average and Median New Unit rent by zipcode",
       x = "",
       y = "") +
  theme_minimal() 
```

**Geographic heatmap showing median rent across zip codes**

```{r}
#Zipcodes 
zip_shapes <- zctas(cb = FALSE)
names(zip_shapes)
```

```{r}
#| dpi: 100

# OLD RENT MAP

# Calculate median rent by zipcode
geo_rent_summary <- old_rents |>
  group_by(original_zip_code) |>
  summarise(average_rent = median(original_unit_rent, na.rm = TRUE)) |>
  mutate(zipcode = as.character(original_zip_code))

# Zipcodes for joining
map_data_old <- zip_shapes |>
  mutate(ZCTA5CE20 = as.character(ZCTA5CE20)) |>
  left_join(geo_rent_summary, by = c("ZCTA5CE20" = "zipcode")) |> 
  filter(ZCTA5CE20 %in% geo_rent_summary$zipcode)

# Plot with Zip labels
p1 <- ggplot(map_data_old) +
  geom_sf(aes(fill = average_rent)) +
  geom_sf_text(aes(label = ZCTA5CE20),
               color = "white",
               size = 2.5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Rent",
    title = "Original Median Rent"
  ) +
  theme_void()



# NEW RENT MAP

# Calculate median rent by zipcode
geo_new_rent <- new_rents |>
  group_by(new_zip_code) |>
  summarise(avg_new_rent = median(rent_amount, na.rm = TRUE)) |>
  mutate(zipcode = as.character(new_zip_code))

map_data_new <- zip_shapes |>
  mutate(ZCTA5CE20 = as.character(ZCTA5CE20)) |>
  left_join(geo_new_rent, by = c("ZCTA5CE20" = "zipcode")) |> 
  filter(ZCTA5CE20 %in% geo_new_rent$zipcode)

# Plot with Zip labels
p2 <- ggplot(map_data_new) +
  geom_sf(aes(fill = avg_new_rent), color = "lightblue") +
  geom_sf_text(aes(label = ZCTA5CE20),
               color = "white",
               size = 2.5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Rent",
    title = "Median Relocation Rent"
  ) +
  theme_void()

#Side by Side  

p1 + p2
```
Most program clients were located in the 20019, 20020, 20032, and 20002 zip codes, specifically in the Northeast and Southeast areas, where rental costs are significantly lower.

**Median rent by unit size overall**

```{r}
# Average rent + number of participants by unit size
median_rent_unit <- new_rents |>
  filter(!is.na(proposed_new_unit_size)) |>
  group_by(proposed_new_unit_size) |>
  summarise(
    n_clients  = n(),                                 # count of participants     
    median_new_rent = median(rent_amount, na.rm = TRUE)    # median rent
  )

median_rent_unit

# If your summary is median_rent_unit
median_rent_unit_long <- median_rent_unit |>
  select(proposed_new_unit_size, median_new_rent,  n_clients) |>
  pivot_longer(
    cols = c(median_new_rent),
    names_to = "rent_type",
    values_to = "rent"
  )
```

```{r}
# Median rent as bar, n_clients as label
ggplot(median_rent_unit,
       aes(x = proposed_new_unit_size,
           y = median_new_rent, fill = proposed_new_unit_size)) +
  geom_col(aes(y = median_new_rent), position = position_dodge(width = 0.9)) +
  geom_text(aes(label = paste0("n = ", n_clients)),
            vjust = -0.3, size = 3) +
  scale_y_continuous(labels = dollar) + 
  labs(
    title = "Median Monthly Rent By Unit Size After Relocation",
    subtitle = "Labels show the number of participants in each unit type",
    x = "",
    y = "",
    fill = "Unit Size"
  ) +
  theme_minimal()
```
Most program clients are concentrated in 2 bedroom (240 clients) and 3 bedroom (162 clients) units. Though larger units attract more rent, there are few program participants in the 4 or 5 bedroom units. Each bedroom is associated with roughly $713 higher rent. 

**Key factors driving rent prices** - **Regression Analysis**

```{r}
# Run linear regression model to quantify how variables such as unit size and location influence rent costs,
relocation_model <- lm(rent_amount ~ proposed_new_unit_size + new_zip_code + household_size + 
                       utilities_included, data = new_rents)

# regression results
summary(relocation_model)
```

About 62% of the variation in rent amounts is explained by unit size and Zipcode.



## Comparing Old Rents and Relocation Rents

```{r}
# Join old and new rents
compare_old_new <- old_rents |> 
  select(response_id, old_rent = original_unit_rent, 
         old_zip = original_zip_code) |> 
  inner_join(
    new_rents |> 
      select(response_id, new_rent = rent_amount, new_zip = new_zip_code,
             household_size, utilities_included),
    by = "response_id"
  ) |> 
  drop_na(old_rent, new_rent) 
head(compare_old_new)

# Check if any NAs remain
any(is.na(compare_old_new))
```


```{r}
#Adding a logical if rent increaased after relocation 
compare_old_new <- compare_old_new |>
  mutate(
    rent_change = new_rent - old_rent,
    rent_increase = rent_change > 0
  )
head(compare_old_new)
```

```{r}
# Summary statistics 
stats <- compare_old_new |> 
  summarise(
    mean_rent = mean(rent_change),
    median_rent = median(rent_change),
    min_rent = min(rent_change),
    max_rent = max(rent_change)
  )
stats
```

The largest decrease in rent was 2,215 meaning some clients actually pay less after relocating. Half of the clients had their rent increase by 825 and on average clients saw a rent increase of about 979 suggesting some rent increases are pulling the average rent up. Largest increase was 3865. In general, most clients see an increase in rent as they move.

```{r}
# Count of increases vs decreases
table(compare_old_new$rent_increase)
prop.table(table(compare_old_new$rent_increase))  # proportion of increases
```

```{r}
# Histogram of rent changes
ggplot(compare_old_new, aes(x = rent_change)) +
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black") +
  labs(title = "Rent changes after relocation",
       x = "Rent Change ($)",
       y = " Count")
```

```{r}
#Propotion of rent increase per zipcode. 
compare_old_new |> 
  group_by(new_zip) |> 
  summarise(
    avg_increase = mean(rent_change),
    prop_increase = mean(rent_increase),
    .groups = "drop"
  ) |> 
  arrange(avg_increase)
```

80% of clients who moved to zipcode 20020 experienced an average of 433.8 increase in rent. All clients who moved to zipcode 6 had an average of 2578 increase in rent.

## Move type - same zipcode or not

```{r}
#Check if clients moved to different zipcode or not
compare_old_new  <- compare_old_new |>
  mutate(
    same_zip = old_zip == new_zip,          # TRUE if zip did not change
    move_type = ifelse(same_zip, "Stayed", "Moved") 
  )
head(compare_old_new)

#table showing count 
table(compare_old_new$move_type)
```

```{r}
ggplot(compare_old_new, aes(x = move_type, fill = move_type)) +
  geom_bar() +
  labs(title = "Clients moving vs staying by zipcode",
       x = "Move Type",
       y = "Count",
       fill = "Move Type") +
  theme_minimal()
```



## Current Rents

**Current Rental Trends**

```{r}
current_rents <- rent[["curr"]]
head(current_rents)

cat("Missing values present:", any(is.na(current_rents)), "\n")
```

```{r}
#change names to snake case 
current_rents <- clean_names(current_rents)

# Check new names
colnames(current_rents)
```

```{r}
# Selecting variables of interest
current_rents <- current_rents |>
  select(response_id, full_rent, tcp_total_incl_adj_credit,
         client_rent_amount, housing_unit_city_state_zip) |>
  mutate(
    # extract zipcode 
    current_zip = str_extract(housing_unit_city_state_zip, "\\d{5}$"),
    # convert  to factor
    current_zip = as.factor(current_zip),
    # rename tcp_total column
    rent_assistance = tcp_total_incl_adj_credit,
    # convert response_id from dbl to factor
    response_id = as.factor(response_id)
  ) |>
  select(response_id, full_rent, rent_assistance,
         client_rent_amount, current_zip) |>
  drop_na()
head(current_rents)
```

**Current Unit Rents Summary Statistics**

```{r}
# Descriptive Statistics for full rent before subsidies 
current_rents |>
  summarize(
    mean = mean(full_rent, na.rm = TRUE),
    median = median(full_rent, na.rm = TRUE),
    min = min(full_rent, na.rm = TRUE),
    max = max(full_rent, na.rm = TRUE),
    sd = sd(full_rent, na.rm = TRUE)
  )


# Descriptive Statistics for client_rent_amount - portion of rent paid by client 
current_rents |>
  summarize(
    mean = mean(client_rent_amount, na.rm = TRUE),
    median = median(client_rent_amount, na.rm = TRUE),
    min = min(client_rent_amount, na.rm = TRUE),
    max = max(client_rent_amount, na.rm = TRUE),
    sd = sd(client_rent_amount, na.rm = TRUE)
  )


# Descriptive Statistics for rent_assistance- portion of rent paid by program
current_rents |>
  summarize(
    mean = mean(rent_assistance, na.rm = TRUE),
    median = median(rent_assistance, na.rm = TRUE),
    min = min(rent_assistance, na.rm = TRUE),
    max = max(rent_assistance, na.rm = TRUE),
    sd = sd(rent_assistance, na.rm = TRUE)
  )
```

**Distribution of current unit rents**

```{r}
# Distribution - full rent 
med_full <- median(current_rents$full_rent, na.rm = TRUE)

ggplot(current_rents, aes(x = full_rent)) +
  geom_histogram(fill = "seagreen", color = "black", bins = 30) +
  geom_vline(aes(xintercept = med_full), color = "red", 
             linetype = "dashed", size = 1) +
  annotate(
    "text",
    x = med_full * 1.03,          # slight shift right
    y = Inf,
    vjust = 1.5,
    label = paste("Median:", dollar(med_full)),
    color = "red",
    size = 4
  ) +
  labs(title = "Distribution of Current Full Rents", 
       subtitle = "Total rent charged for each unit before applying any subsidies.",
       x = "", 
       y = "") +
  scale_x_continuous(labels = dollar) +
  theme_minimal()
```


**Breakdown of full rent, rent assistance, and client share**

```{r}
# Summarize median and max for each rent type
current_rent_summary <- current_rents |> 
  summarize(
    full_rent_med = median(full_rent, na.rm = TRUE),
    full_rent_max = max(full_rent, na.rm = TRUE),
    client_rent_amount_med = median(client_rent_amount, na.rm = TRUE),
    client_rent_amount_max = max(client_rent_amount, na.rm = TRUE),
    rent_assistance_med = median(rent_assistance, na.rm = TRUE),
    rent_assistance_max = max(rent_assistance, na.rm = TRUE)
  ) |> 
  
  # Convert to long format for ggplot
  pivot_longer(
    cols = everything(),
    names_to = c("rent_type", "stat"),
    names_pattern = "(.*)_(med|max)$",
    values_to = "amount"
  ) |> 
  mutate(
    # Rename statistic labels
    stat = case_when(
      stat == "med" ~ "Median",
      stat == "max" ~ "Max"
    ),
    # Rename rent types
    rent_type = case_when(
      rent_type == "full_rent" ~ "Full Rent",
      rent_type == "client_rent_amount" ~ "Client Rent",
      rent_type == "rent_assistance" ~ "Rent Assistance"
    )
  )

# Plot 
ggplot(current_rent_summary, aes(x = stat, y = amount, fill = rent_type)) +
  geom_col(position = position_dodge(width = 0.8), color = "black") +  # Bars side by side
  geom_text(
    aes(label = scales::label_number(accuracy = 1, prefix = "$")(amount)), #int numbers
    position = position_dodge(width = 0.8),
    vjust = -0.5,
    size = 3
  ) +  
  scale_fill_manual(values = c("Full Rent" = "seagreen",
                               "Client Rent" = "orange",
                               "Rent Assistance" = "skyblue")) +
  scale_y_continuous(labels = scales::label_number(accuracy = 1, prefix = "$")) +
  labs(
    title = "Comparison of Current Rent Amounts",
    subtitle = "Median and Max for Full Rent, Client Rent and Rent Assistance",
    x = "",
    y = "",
    fill = "Rent Type"
  ) +
  theme_minimal()

```


```{r}
#Number of clients by current Zip Code 
# Current zipcodes by number of program clients
current_zipcodes <- current_rents|>
  filter(!is.na(current_zip)) |>
  group_by(current_zip) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count))

# Top 5 new zipcodes
top_curr_zip <- current_zipcodes |> 
  filter(row_number() <= 5) |> 
  pull(current_zip) |> 
  sort()

# Plot
ggplot(current_zipcodes, aes(x = reorder(current_zip, count), y = count)) +
  geom_bar(stat = "identity", fill = "purple") +
  geom_text(aes(label = count), hjust = -0.1) + # Count labels outside bars
  scale_y_continuous(position = "right") +
  coord_flip() +  # horizontal bars
  labs(
    title = "Number of clients by current zip code",
    subtitle = paste("Top 5 zip codes:", paste(top_curr_zip, collapse = ", ")),
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```


**Rent distribution at different stages**

```{r}
# Descriptive Statistics and a Box plot to compare distribution of rents old, relocation and current rent stages
# Convert response_id to character
old_rents <- old_rents |> mutate(response_id = as.character(response_id))
new_rents <- new_rents |> mutate(response_id = as.character(response_id))
current_rents <- current_rents |> mutate(response_id = as.character(response_id))

# Combine all rent stages for comparison and add rent_type
all_rents <- bind_rows(
  old_rents |> select(response_id, rent = original_unit_rent) |> mutate(rent_type = "old"),
  new_rents |> select(response_id, rent = rent_amount) |> mutate(rent_type = "new"),
  current_rents |> select(response_id, rent = full_rent) |> mutate(rent_type = "curr")
) |> mutate(rent_type = factor(rent_type, levels = c("old", "new", "curr")))


# Visualization: Boxplot of Rents by Stage
ggplot(all_rents, aes(x = rent_type, y = rent, fill = rent_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.1, width = 0.2) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Distribution of Rents by Stage",
    x = "",
    y = "",
    fill = "Rent Stage"
  ) +
  theme_minimal()
```

**Current Average Rent vs Market Average Rent**

```{r}
zillow_rent <- tibble::tribble(
  ~zip_code, ~average_rent,
  "20001", "$2,695",
  "20002", "$2,527",
  "20003", "$2,744",
  "20005", "$2,750",
  "20007", "$3,400",
  "20008", "$2,300",
  "20009", "$2,700",
  "20010", "$2,369",
  "20011", "$2,443",
  "20012", "$2,290",
  "20016", "$2,622",
  "20017", "$1,750",
  "20018", "$2,250",
  "20019", "$2,100",
  "20020", "$2,233",
  "20024", "$2,378",
  "20032", "$1,550",
  "20037", "$2,600"
) |>
  mutate(
    zip_code = as.character(zip_code),
    zillow_rent = as.numeric(str_remove_all(average_rent, "[$,\\.]"))
  ) |>
  select(zip_code, zillow_rent)
```

```{r}
# Average and media rent by zipcode
current_zip_rent <- current_rents |>
  group_by(current_zip) |>
  summarize(average_rent = mean(full_rent, na.rm = TRUE),
            median_rent = median(full_rent, na.rm = TRUE)) |>
  arrange(average_rent)
current_zip_rent
```

```{r}
#Current average full rents vs Zillow average rent 
# Current rents
current_zip_rent_summ <- current_zip_rent |>
  group_by(current_zip) |>
  summarise(
    average_rent = mean(average_rent, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(zip_code = as.character(current_zip)) |>
  select(zip_code, average_rent)

# Join current and zillow
current_zillow_top10 <- current_zip_rent_summ |>
  inner_join(zillow_rent, by = "zip_code") |>
  slice_max(order_by = average_rent, n = 10)

# Long format for plotting
rents_long <- current_zillow_top10 |>
  pivot_longer(
    cols = c(average_rent, zillow_rent),
    names_to = "rent_type",
    values_to = "rent"
  ) |>
  mutate(
    rent_type = case_when(
      rent_type == "average_rent" ~ "Current average rent",
      rent_type == "zillow_rent"  ~ "Zillow average rent"
    )
  )

```

```{r}
# Plot
ggplot(rents_long, aes(x = zip_code, y = rent, fill = rent_type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7) +
  scale_fill_manual(
    values = c(
      "Current average rent" = "steelblue",
      "Zillow average rent"  = "maroon"
    )
  ) +
  scale_y_continuous(position = "right", labels = scales::dollar) +
  coord_flip() +
  labs(
    title = "Current vs Zillow Average Rent by Zipcode",
    subtitle = "Top 10 client zip codes by current average rent",
    x = "",
    y = "",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 1),
    legend.position = "top"
  )
```



## Current Rent Comparison with Fair Market Rent 

```{r}
#API tokens - USHUD 
#Token expires and needs to be updated periodically. Here is the website to get new token:https://www.huduser.gov/hudapi/public/save-selection
hud_token <- "eyJ0eXAiOiJKV1QiLCJhbGciOiJSUzI1NiJ9.eyJhdWQiOiI2IiwianRpIjoiMTZjMWNmM2IzYzA5NjdiMjk5ODVjOTc3ZDE0ZDIxNWM0ZmU2MzEwMGM2NjU1NTkwOTY3ZTFmZWE4MzM2NmNiMDE5YWQyNWYyOTA5ZjNlY2EiLCJpYXQiOjE3NjE0NTQ3MDUuNTk0ODM5LCJuYmYiOjE3NjE0NTQ3MDUuNTk0ODQzLCJleHAiOjIwNzY5ODc1MDUuNTg5MjU5LCJzdWIiOiIxMTAxMjMiLCJzY29wZXMiOltdfQ.XZ4VYcUnpx9mX9ofU59CYDrZtVVXRggkIrpM4JI4lhtZUlB6SwRcJ93IY2fU1HbxSR_jSU_9tTr67obW3mTbLQ"
```

```{r}
# Send a GET request to HUD API with token
test <- GET(
  "https://www.huduser.gov/hudapi/public/fmr/listStates",
  add_headers(Authorization = paste("Bearer", hud_token))
)

# Confirm request success 
status_code(test)
output <- content(test)
```

```{r}
#URL and Authorization Header
hud_url <- "https://www.huduser.gov/hudapi/public/fmr"
hud_states <- paste0(hud_url, "/listStates")
hud_dc <- paste0(hud_url, "/statedata/DC")
hud_dc_metro <- paste0(hud_url, "/data/METRO47900M47900")
auth_header <- paste("Bearer", hud_token)

```

```{r}
# Get list of available states
req_states <- GET(hud_states, add_headers(Authorization = auth_header))
status_code(req_states)
```

```{r}
#Washington DC Metro Housing rentals data 
req_dc <- GET(url = hud_dc, add_headers(Authorization = auth_header))
df_dc <- NULL
dc_metro_code <- NULL

if (status_code(req_dc) == 200) {
  df_dc <- content(req_dc) 
  df_dc <- df_dc$data$metroareas |> 
    map_dfr(function(d) as.data.frame(d))
  
  dc_metro_code <- df_dc$code[1]
}

dc_metro_code
df_dc
```

```{r}
#DC API basic rent data 
req_dc_metro <- GET(url = paste0(hud_dc_metro, "?updated=2020"), 
                    add_headers(Authorization = auth_header))
df_dc_metro <- NULL

if (status_code(req_dc_metro) == 200) {
  df_dc_metro <- content(req_dc_metro) 
  df_dc_metro <- df_dc_metro$data$basicdata |> 
    map_dfr(function(d) as.data.frame(d))
  
}

#inspect rows
head(df_dc_metro)
```

```{r}
# Check for missing values
colSums(is.na(df_dc_metro))
```

```{r}
#Clean names 
df_dc_metro<- clean_names(df_dc_metro)

# Check new names
colnames(df_dc_metro)
```

```{r}
#Descriptive Stats 
df_dc_metro |>
  select(efficiency, one_bedroom, two_bedroom, three_bedroom, four_bedroom) |>
  map(summary)
```

```{r}
# Create avg.rent after removing "MSA level"
df_dc_metro <- df_dc_metro |> 
  filter(zip_code != "MSA level") |> 
  rowwise() |> 
  mutate(avg.rent = median(c(efficiency, one_bedroom, two_bedroom, 
                             three_bedroom, four_bedroom), na.rm = TRUE)) |> 
  ungroup()

# Get top 10 most expensive zip codes
top_rent_zips <- df_dc_metro |>
  arrange(zip_code) |> 
  slice_head(n = 10)
  
top_rent_zips
```

```{r}
# Merging relocation rents and market rents
# Reshape to long format: one row per zip and bedroom size
hud_rents <- df_dc_metro |>
  select(zip_code, efficiency, one_bedroom, two_bedroom, 
         three_bedroom, four_bedroom) |>
  pivot_longer(
    cols = c(efficiency, one_bedroom, two_bedroom, three_bedroom, four_bedroom),
    names_to = "bedroom_type",
    values_to = "market_rent"
  ) |>
  mutate(
    bedrooms = case_when(
      bedroom_type == "efficiency" ~ 0,
      bedroom_type == "one_bedroom" ~ 1,
      bedroom_type == "two_bedroom" ~ 2,
      bedroom_type == "three_bedroom" ~ 3,
      bedroom_type == "four_bedroom" ~ 4,
      TRUE ~ NA_integer_
    )
  ) |>
select(-bedroom_type) |>
  arrange(desc(market_rent))

head(hud_rents)
```


**Fair Market Rents Summary Statistics**

```{r}
# HUD summary statistics
hud_rents |>
  summarize(
    mean = mean(market_rent, na.rm = TRUE),
    median = median(market_rent, na.rm = TRUE),
    min = min(market_rent, na.rm = TRUE),
    max = max(market_rent, na.rm = TRUE),
    sd = sd(market_rent, na.rm = TRUE)
  )

```

```{r}
# Get zipcodes present in both datasets
common_zips <- intersect(unique(new_rents$new_zip_code), 
                         unique(hud_rents$zip_code))

# Filter both datasets to only these common zipcodes
new_rents <- new_rents |> 
  filter(new_zip_code %in% common_zips)

hud_rents <- hud_rents |> 
  filter(zip_code %in% common_zips)

new_rents 
hud_rents
```

**HUD fair market rents by unit size**

```{r, echo=TRUE}
# Market median rent by unit size 
market_median<- hud_rents |>
  group_by(bedrooms) |>
  filter(!is.na(bedrooms)) |>
  summarise(median_mkt_rent = median(market_rent, na.rm = TRUE))

ggplot(market_median, aes(x = bedrooms, fill = bedrooms)) +
  geom_col(aes(y = median_mkt_rent), position = position_dodge(width = 0.9)) +
  geom_text(aes(y = median_mkt_rent, label = dollar(round(median_mkt_rent, 0))),
            vjust = -0.5, size = 3.5) +  # Labels on top of bars
  scale_y_continuous(labels = dollar) + 
  labs(title = "FMR Rents by Unit Size",
       subtitle = "Roughly 40% of renters pay this amount or less", 
       x = "",
       y = "",
       fill = "Unit Size") +
  theme_minimal()
```

```{r}
# Combine all rent stages for comparison and add rent_type
combined_rents <- old_rents |>
  select(original_unit_rent, original_zip_code) |> 
  # Full join with new rents on zip code, allow many-to-many relationship 
  full_join(
    new_rents |>  
      select(rent_amount, new_zip_code, proposed_new_unit_size), 
    by = c("original_zip_code" = "new_zip_code"), 
    relationship = "many-to-many"
  ) |> 
  # Full join with current rents on zip code, allow many-to-many relationship
  full_join(
    current_rents |>  
      select(full_rent, client_rent_amount, rent_assistance, current_zip), 
    by = c("original_zip_code" = "current_zip"),
    relationship = "many-to-many"
  ) |> 
  mutate(
    rent_old = original_unit_rent,                   # Rename old rent variable 
    rent_new = rent_amount,                          # Rename new rent variable
    bedrooms = as.integer(str_extract(proposed_new_unit_size, "\\d+")),  
    zip_code = original_zip_code                     # Use original zip code consistently 
  ) |> 
  group_by(zip_code, bedrooms) |>            # Group data by zip code and number of bedrooms
  summarise(
    median_rent_old = round(median(rent_old, na.rm = TRUE)),              
    median_rent_new = round(median(rent_new, na.rm = TRUE)),            
    median_full_rent = round(median(full_rent, na.rm = TRUE)),            
    median_client_rent = round(median(client_rent_amount, na.rm = TRUE)), 
    median_rent_assistance = round(median(rent_assistance, na.rm = TRUE)),
    .groups = "drop"                                                     # Ungroup after summarise
  )

head(combined_rents)

```

```{r}
# Join with all datasets by response_id with many-to-many relationship
combined_rents <- old_rents |>
  select(response_id, original_unit_rent, original_zip_code) |> 
  full_join(
    new_rents |>  
      select(response_id, rent_amount, new_zip_code, proposed_new_unit_size), 
    by = "response_id", 
    relationship = "many-to-many"
  ) |> 
  full_join(
    current_rents |>  
      select(response_id, full_rent, client_rent_amount, rent_assistance, current_zip), 
    by = "response_id",
    relationship = "many-to-many"
  ) |> 
  mutate(
    rent_old = original_unit_rent,
    rent_new = rent_amount,
    zip_code = coalesce(new_zip_code, current_zip, original_zip_code),
    bedrooms = as.integer(str_extract(proposed_new_unit_size, "\\d+"))
  )

```

```{r}
# Join with HUD rents by zip_code and bedrooms 
rent_map <- combined_rents |> 
  left_join(hud_rents, by = c("zip_code", "bedrooms"), relationship = "many-to-many")

# Impute missing rent and assistance values 
rent_map <- rent_map |> 
  group_by(zip_code, bedrooms) |> 
  mutate(
    rent_old = ifelse(is.na(rent_old), median(rent_old, na.rm = TRUE), rent_old),                 # Impute missing rent_old
    rent_new = ifelse(is.na(rent_new), median(rent_new, na.rm = TRUE), rent_new),                 # Impute missing rent_new
    full_rent = ifelse(is.na(full_rent), median(full_rent, na.rm = TRUE), full_rent),             # Impute missing full_rent
    client_rent_amount = ifelse(is.na(client_rent_amount), median(client_rent_amount, na.rm = TRUE), client_rent_amount),  # Impute missing client_rent_amount
    rent_assistance = ifelse(is.na(rent_assistance), median(rent_assistance, na.rm = TRUE), rent_assistance),              # Impute missing rent_assistance
    market_rent = ifelse(is.na(market_rent), median(market_rent, na.rm = TRUE), market_rent)      # Impute missing market_rent (HUD)
  ) |> 
  ungroup()

# Drop rows missing bedrooms because its a key variable
rent_map <- rent_map |> filter(!is.na(bedrooms)) 
# Group by zip_code and bedrooms and summarize rents by median, rounded to whole dollars
rent_summary <- rent_map |> 
  group_by(zip_code, bedrooms) |>
  summarise(
    median_rent_old = round(median(rent_old, na.rm = TRUE)),
    median_rent_new = round(median(rent_new, na.rm = TRUE)),
    median_full_rent = round(median(full_rent, na.rm = TRUE)),
    median_client_rent = round(median(client_rent_amount, na.rm = TRUE)),
    median_rent_assistance = round(median(rent_assistance, na.rm = TRUE)),
    median_hud_rent = round(median(market_rent, na.rm = TRUE)),
    .groups = "drop"
  ) 

rent_summary
```

```{r}
# Summarize average rents by zipcode and bedroom count. Only current rent was used. 
avg_rents <- rent_map |>
  group_by(zip_code, bedrooms) |>
  summarize(
    avg_current_rent = median(full_rent, na.rm = TRUE),
    avg_market_rent = median(market_rent, na.rm = TRUE),
    .groups = "drop"
  )

# Identify top 6 zip codes by average relocation rent overall
top6_zips <- avg_rents |>
  group_by(zip_code) |>
  summarize(overall_avg = median(avg_current_rent, na.rm = TRUE)) |>
  slice_max(order_by = overall_avg, n = 6) |> 
  pull(zip_code)


# Filter for top 6 zips
top6_zips_plot <- avg_rents |> 
  filter(zip_code %in% top6_zips) |> 
  pivot_longer(cols = c(avg_current_rent, avg_market_rent), 
               names_to = "rent_type", values_to = "avg_rent")

# Plot grouped bar chart with flipped coordinates
ggplot(top6_zips_plot, aes(x = factor(bedrooms), y = avg_rent, 
                           fill = rent_type)) +
  geom_col(position = position_dodge(width = 0.7)) +
  facet_wrap(~ factor(zip_code), scales = "free_y") +
  coord_flip() +
  scale_fill_manual(
    values = c(avg_current_rent = "steelblue", avg_market_rent = "tomato"),
    labels = c("Career MAP Rent", "Market Rent")
  ) +
  labs(
    title = "Median Current and FMR Rents by Zip code and Beds",
    x = "Bedroom",
    y = "Amounts ($)",
    fill = "Rent Type"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```


## DC Housing Authority 2025 Maximum Housing Payment Standards 
**Maximum unit rents in listed zip codes**

```{r}
# Max payment standard per unit size (0–5 bedrooms)
unit_sizes <- sort(unique(rent_summary$bedrooms))  # get unit sizes present
max_payment_std <- c(3762, 3845, 4327, 5410, 6382, 7340)  

max_payment_df <- tibble(
  bedrooms = unit_sizes,
  rent_type = "Max Payment Standard",
  rent = max_payment_std[1:length(unit_sizes)]
)
```

```{r}
rent_compare_df <- rent_summary |> 
  group_by(bedrooms) |> 
  summarise(
    "Current Median Rent" = median(median_full_rent, na.rm = TRUE),
    "HUD Median Rents" = median(median_hud_rent, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  pivot_longer(
    cols = c("Current Median Rent", "HUD Median Rents"),
    names_to = "rent_type",
    values_to = "rent"
  )
```


```{r, include=FALSE}
all_rents_long <- bind_rows(rent_compare_df, max_payment_df)

```

```{r, warning=FALSE}
#Plot 
ggplot(all_rents_long, aes(x = factor(bedrooms), y = rent, fill = rent_type)) +
  geom_col(position = position_dodge(width = 0.4), width = 0.4) +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    title = "Median Rent Comparison by Unit Size",
    subtitle = "Current Rents vs HUD Rents vs Max Payment Standard",
    x = "Unit Sizes",
    y = "",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "top")

```

```{r, include=FALSE}
#Inflation Adjusted Predictions 
dc_inflation_rate <- 0.025   # 2.5% annual inflation
years_ahead <- 1            # project 1 year forward

# Adjusted predictions
rent_forecast <- current_zillow_top10 |>
  mutate(
    predicted_current_rent = average_rent * (1 + dc_inflation_rate)^years_ahead,
    predicted_zillow_rent = zillow_rent * (1 + dc_inflation_rate)^years_ahead
  )

# Keep only top 5 ZIP codes by current average rent
top5_rent_forecast <- rent_forecast |>
  slice_max(order_by = average_rent, n = 5)

```

```{r}
# Keep only top 5 ZIP codes by current average rent
top5_rent_forecast <- rent_forecast |>
  slice_max(order_by = average_rent, n = 5)

#Convert to long format 
rents_long_forecast <-top5_rent_forecast|>
  select(zip_code, average_rent, zillow_rent,
         predicted_current_rent, predicted_zillow_rent) |>
  pivot_longer(
    cols = -zip_code,
    names_to = "rent_type",
    values_to = "rent"
  ) |>
  mutate(
    rent_type = dplyr::recode(
      rent_type,
      average_rent = "Current Avg Rent",
      zillow_rent = "Zillow Avg Rent",
      predicted_current_rent = "Predicted Rent",
      predicted_zillow_rent = "Predicted Zillow Rent"
    )
  )

```

Annual Inflation rate for DC area as of September 2025 was 2.5%. 
```{r}
#Plot rent vs inflation adjusted rents 
ggplot(rents_long_forecast, aes(x = zip_code, y = rent, fill = rent_type)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  scale_y_continuous(labels = scales::dollar) +
  coord_flip() +
  labs(
    title = "Current and Predicted Rent by Zip Code",
    subtitle = "Inflation-adjusted projection using 2.5 % DC annual inflation rate",
    x = "",
    y = "",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "top")

```

```{r}
#Projection table
rent_forecast |>
  select(zip_code,
         average_rent,
         predicted_current_rent,
         zillow_rent,
         predicted_zillow_rent) |>
  mutate(across(where(is.numeric), round, 0))
rent_forecast 
```





